

Recommendation system is used in day to day life. It is used in book search, online shopping, movie search, social networking, to name a few. Recommendation system applies statistical and knowledge discovery techniques to provide recommendation to new item to the user based on previously recorded data. The recommendation information can be used to increase customer retention, promote cross-selling, and add value to buyer-seller relationship. 

Broadly recommender systems are classified into two categories:
  - Content based: recommending items that shares some common attributes based on user preferences
  - Collaborative filtering: recommending item from users sharing common preferences.
  
Commonly used metrics to quantify the performace of recommender systems are Root Mean Squared Error (RMSE), precision and Recall. 

R has a nice package recommenderlab that provides infrastructure to develop and test recommender algorithm. recommenderlab focusses on recommender algorithm based on collaborative filtering.

I used **recommenderlab** to get insight into collaborative filtering algorithms and evalaute the performace of different algorithm available in the framework on Movie Lens 100k dataset. The dataset is downloaded from [here](http://files.grouplens.org/datasets/movielens/ml-100k/).



```{r echo=FALSE, warning=FALSE, message=FALSE}
###Recommender System algorithm implementaion on Movie Lens 100k data ###

## load libraries ####
library(recommenderlab)
library(reshape2)

#Function Definitions ##

#### Read Data ####
## data downloaded from http://grouplens.org/datasets/movielens/ 

## read the rating data for all users 
readData <- function(){
  
  ratingDF <- read.delim("./data/u.data", header=F)
  colnames(ratingDF) <- c("userID","movieID","rating", "timestamp")
  
  ## read movie data
  moviesDF <- read.delim("./data/u.item", sep="|", header=F, stringsAsFactors = FALSE)
  colnames(moviesDF)[colnames(moviesDF)=="V1"] <- "movieID"
  colnames(moviesDF)[colnames(moviesDF)=="V2"] <- "name"
  
  return(list(ratingDF=ratingDF, movieDF=moviesDF)) 
}

#### data Cleansing and processing ####
preProcess = function(ratingDF, moviesDF)
{ 
  ratingDF[,2] <- dataList$movieDF$name[as.numeric(ratingDF[,2])]
  
  # remove duplicate entries for any user-movie combination
  ratingDF <- ratingDF[!duplicated(ratingDF[,1:2]),]
}

## Create movie ratingMatrix from rating Data and movie data ####
createRatingMatrix <- function(ratingDF)
{
  # converting the ratingData data frame into rating marix
  ratingDF_tmp <- dcast( ratingDF, userID ~ movieID, value.var = "rating" , index="userID")
  ratingDF <- ratingDF_tmp[,2:ncol(ratingDF_tmp)]
  
  ratingMat <- as(ratingDF, "matrix")  ## cast data frame as matrix
  movieRatingMat <- as(ratingMat, "realRatingMatrix")   ## create the realRatingMatrix
  ### setting up the dimnames ###
  dimnames(movieRatingMat)[[1]] <- row.names(ratingDF)
  return (movieRatingMat)
}


##### Create Recommender Model ####
evaluateModels <- function(movieRatingMat)
{
  
  ## Find out and anlayse available  recommendation algorithm option for realRatingMatrix data
  recommenderRegistry$get_entries(dataType = "realRatingMatrix")
  
  scheme <- evaluationScheme(movieRatingMat, method = "split", train = .9,
                             k = 1, given = 10, goodRating = 4)
  
  algorithms <- list(
    RANDOM = list(name="RANDOM", param=NULL),
    POPULAR = list(name="POPULAR", param=NULL),
    UBCF = list(name="UBCF", param=NULL)
  )
  
  # run algorithms, predict next n movies
  results <- evaluate(scheme, algorithms, n=c(1, 3, 5, 10, 15, 20))
  
  ## select the first results
  
  return (results)
}


visualise <- function(results)
{
  # Draw ROC curve
  plot(results, annotate = 1:3, legend="topright")
  
  # See precision / recall
  plot(results, "prec/rec", annotate=3, legend="topright", xlim=c(0,.22))
}


#### Create prediction model ####
createModel <-function (movieRatingMat,method){
  
  model <- Recommender(movieRatingMat, method = method)
  names(getModel(model))
  getModel(model)$method
  
  getModel(model)$nn
  
  return (model)
}


### Predict user rating using UBCF recommendation algoithm ####
recommendations <- function(movieRatingMat, model, userID, n)
{
  
  ### predict top n recommendations for given user
  topN_recommendList <-predict(model,movieRatingMat[userID],n=n)
  as(topN_recommendList,"list")
}

```



```{r}
###### Recommender System algorithm implementaion on Movie Lens 100k data ###

## load libraries ####
library(recommenderlab)
library(reshape2)


# Load Movie Lens data
dataList<- readData()
# data cleansing and preprocessing
ratingDF<- preProcess(dataList$ratingDF, dataList$movieDF)
# create movie rating matrix
movieRatingMat<- createRatingMatrix(ratingDF)
# evaluate models
evalList <- evaluateModels(movieRatingMat)
```

The plot for comparing "Random", "Popular", "UBCF" based recommender algorithm is shown:

```{r fig.width=7, fig.height=6}
# plot evaluation result
visualise(evalList)
```

The visualisation shows "UBCF" algorithm has highest precision. So I picked "UBCF" to predicts top 10  recommendation of user with userID = 1. 

```{r fig.width=7, fig.height=6}
## on visualization, looks like UBCF has highest precision.
# get Confusion matrix for "UBCF"
getConfusionMatrix(evalList[["UBCF"]])[[1]][,1:4]

## run "UBCF" recommender
rec_model <- createModel(movieRatingMat, "UBCF")
userID <- 1
topN <- 5
recommendations(movieRatingMat, rec_model, userID, topN)
```

The complete R code can be found [here]().